Edge sign prediction in general graphs and connection with Correlation Clustering

### 2017-07-05 gtx_algorithm.tex
add references to Alon '95 paper
first version of the proof of logarithmic iteration of gtx

To achieve our objective of building a spanning tree that minimizes the distances between all
connected pair of nodes in the original graph, we rely on a particular subgraph structure, namely
the star.  We therefore introduce two algorithmic primitives: \extractStar{}, which partition a
graph $G$ into a set of disjoint stars; and \collapseStar{}, which selects edges from $E$ to
assemble these stars into a new, smaller graph. Given a graph topology $G_0=(V_0, E_0)$ and assuming
for simplicity that $G_0$ consists of a single connected component,\footnote{For we can otherwise
run our algorithm in parallel on each connected components of $G_0$.} the \gtx{} algorithm
repeatedly applies these two primitives to produce a sequence of graphs $\{G_t\}_{t=0}^K$ of
decreasing size, until $G_K$ is made of a single node. All the edges selected while reaching
this stage then form the spanning tree we were looking for.
% [from a CMU NIPS submission http://www.stat.cmu.edu/~arinaldo/papers/mutualfriends.pdf] We
% consider Mutual-Friends to be *an algorithmic primitive*, by which we mean a kind of subroutine for
% a more complicated function that iterates Mutual-Friends, similarly to [5].

@@

While all $G_t$ are connected according to \autoref{lem:gtx_stay_connected}, it might happen during
the execution of \extractStar{} that%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_singleton_tikz.pdf}
  \caption{The formation of a singleton star}
  \label{fig:gtx_singleton}
\end{marginfigure}
the degree of a node $u$ drops to zero because all of its neighbors were claimed by the
periphery of previous stars. Such a node then forms a \emph{singleton star}, as illustrated in
\autoref{fig:gtx_singleton}, where after the creation of \starone{1} and \starone{2}, $c_3$ is the
single node of the third star \starone{3}.
In the case no singleton are formed during one execution of \extractStar{}, the number of nodes is
reduced by at least a factor $2$ (because each star is made of at least two nodes). On the other hand, we can
construct a graph with singletons where the factor of reduction in number of nodes and edges can be
arbitrarily close to $1$. Consider \autoref{fig:gtx_noreduc}, where $c_1$ and $c_2$ have a degree
$p$ and their neighbors (in light gray) all have a degree $q$, which we can vary from $1$ to $p-1$.
At this step $t$, we thus have $|V_t| = 2(1+p)+pq$ and $|E_t| = 2(p+qp)$. Next we run \extractStar{}
and assume that we first extract the two stars centered in $c_1$ and $c_2$. This reduces the
effective degree of all the middle nodes (in dark gray) to $0$ and they thus form singletons, that
are all connected to the first two stars by \collapseStar{}. As a result, at step $t+1$ we have%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_noreduc_tikz.pdf}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_noreduc}
\end{marginfigure}
$|V_{t+1}| = 2 + pq$ and $|E_{t+1}| = 2pq$, so that
\begin{equation*}
\frac{|V_{t+1}|}{|V_t|} = \frac{pq+2}{p(q+2) + 2} \qquad \text{and} \qquad
\frac{|E_{t+1}|}{|E_t|} = \frac{2pq}{2p(q+1)} = \frac{q}{q+1}
\end{equation*}
When we let $p$ goes to infinity, setting $q=1$ makes $\frac{|V_{t+1}|}{|V_t|}$ tend to
$\nicefrac{1}{3}$ and $\frac{|E_{t+1}|}{|E_t|}$ to $\nicefrac{1}{2}$ while setting $q=p-1$ makes
both $\frac{|V_{t+1}|}{|V_t|}$ and $\frac{|E_{t+1}|}{|E_t|}$ tend to $1$. Note however that in both
cases, this is not really a problem because at the next iteration, there will be only two stars (one
of them will be a singleton). Is it always like that? In other words, can we have a situation where
a large number of singletons is carried over several consecutive iterations?

>>>>

To achieve our objective of building a spanning tree that minimizes the distances between all
connected pair of nodes in the original graph, we rely on a particular subgraph structure, namely
the star.  We therefore introduce two algorithmic primitives: \extractStar{}, which partition a
graph $G$ into a set of disjoint stars; and \collapseStar{}, which selects edges from $E$ to
assemble these stars into a new, smaller graph. Given a graph topology $G_0=(V_0, E_0)$ and assuming
for simplicity that $G_0$ consists of a single connected component,\footnote{For we can otherwise
run our algorithm in parallel on each connected components of $G_0$.} the \gtx{} algorithm
repeatedly applies these two primitives to produce a sequence of graphs $\{G_t\}_{t=0}^K$ of
decreasing size, until $G_K$ is made of a single node. All the edges selected while reaching
this stage then form the spanning tree we were looking for.
This can be seen as a simple instantiation of a general procedure described in \autocite[Section
5.2]{LowerBound95}.
% [from a CMU NIPS submission http://www.stat.cmu.edu/~arinaldo/papers/mutualfriends.pdf] We
% consider Mutual-Friends to be *an algorithmic primitive*, by which we mean a kind of subroutine for
% a more complicated function that iterates Mutual-Friends, similarly to [5].

@@

While all $G_t$ are connected according to \autoref{lem:gtx_stay_connected}, it might happen during
the execution of \extractStar{} that%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_singleton_tikz.pdf}
  \caption{The formation of a singleton star}
  \label{fig:gtx_singleton}
\end{marginfigure}
the degree of a node $u$ drops to zero because all of its neighbors were claimed by the
periphery of previous stars. Such a node then forms a \emph{singleton star}, as illustrated in
\autoref{fig:gtx_singleton}, where after the creation of \starone{1} and \starone{2}, $c_3$ is the
single node of the third star \starone{3}.

Such singleton stars are problematic because when too many are created, we can end up having
$|V_{t+1}|$ almost as large as $|V_t|$. Consider for instance the graph $G_t$ at the top of
\autoref{fig:gtx_manysingles} (which is actually a tree). It has of order $k^3$ nodes and because
all nodes at level 2 will be selected as center of their star, the graph $G_{t+1}$ (showed at the
bottom of \autoref{fig:gtx_manysingles}) also have around $k^3$ nodes, meaning that
$\frac{|V_{t+1}|}{|V_t|}$ tends to $1$ as $k$ goes to infinity. This is because
$k\left(k(k+1)\right)$ singleton stars were created. Let us denote the fraction of singleton stars
over the total number of nodes by $\alpha$. In that case,
$$\alpha = \frac{k\left(k(k+1)\right)}{1+k+k(k+1)+k^2(k+1)}
= 1 - \frac{1+2k+k^2}{1+2k+2k^2+k^3}
= 1 - \frac{1+k}{1+k+k^2}$$
We will now see that when so many singleton stars are created at one iteration, the next iteration
greatly reduces the number of nodes so that over the course of two iterations, we have the following
result:

\begin{lemma}
  \label{lemma:gtx_lessnodes}
  During the execution of the \gtx{} algorithm, for all $t \leq K-2$, $|V_{t+2}| \leq
  \frac{2}{3}|V_t|$.
\end{lemma}
\begin{marginfigure}
  \centering
  \includegraphics[width=\linewidth]{assets/raw/lots_of_singletons_bw.png}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_manysingles}
\end{marginfigure}
\begin{proof}
  Say that at iteration $t$ of \gtx{}, we have $n_t$ nodes and assume a fraction $\alpha$ of them
  are gonna form singleton stars. This means there will be at most $\frac{1-\alpha}{2}n_t$
  \emph{proper} stars, because such stars are made of at least two nodes. Now fix a constant
  $\lambda \in (0,1)$, for instance $\lambda = \nicefrac{9}{10}$. If $\alpha \leq \lambda$, $n_{t+1}
  \leq (\frac{1-\alpha}{2} + \alpha)n_t = \frac{1+\alpha}{2}n_t = \frac{19}{20}n_t$ and we get a
  constant rate reduction of the number of nodes. On the other hand, if $\alpha > \lambda$, we could
  be in the situation of the previous example and have $\alpha$ tends to $1$. Then we only have a
  few proper stars and by construction, each singleton star is connected to least one proper star.
  In the worst case, those connections are uniform so that each of these proper star has a degree
  $\frac{\alpha n_t}{\frac{1-\alpha}{2}n_t} = \frac{2\alpha}{1-\alpha}$ which goes to infinity when
  $\alpha$ goes to $1$ and is greater than $18$ when $\alpha$ is greater than $\nicefrac{9}{10}$.
  Even ignoring the fact that those proper stars are connected, it follows that in the next
  iteration, the number of node will be reduced by the degree of those proper stars. Setting in the
  worst case $n_{t+1}=\frac{1+\alpha}{2}n_t$, we thus have $$n_{t+2} \leq
  \frac{n_{t+1}}{\frac{2\alpha}{1-\alpha}} = \frac{1-\alpha^2}{4\alpha}n_t$$
  \begin{marginfigure}
    \centering
    \includegraphics[width=\linewidth]{assets/tmp-code/gtx_reduc.pdf}
    \caption{The fraction of $n_t$ nodes remaining in $n_{t+2}$ for the three cases described in the
    main text.}
    % \label{fig:gtx_manysingles}
  \end{marginfigure}
  This defines three ways to go from $n_t$ to $n_{t+2}$: if the first iteration has $\alpha>\lambda$,
  then we just seen that in the worst case we have at least $n_{t+2}\leq
  \frac{1-\lambda^2}{4\lambda}n_t$; if the first iteration has $\alpha < \lambda$, then either the
  second also has  $\alpha < \lambda$ (in which case $n_{t+2}\leq
  \left(\frac{1+\lambda}{2}\right)^2n_t$) or the second has $\alpha>\lambda$ (in which case
  $n_{t+2}\leq \frac{1+\lambda}{2}n_t$). We can optimize over $\lambda$ by finding its value such
  that $\frac{1-\lambda^2}{4\lambda}=\frac{1+\lambda}{2}$, yielding $\lambda = \frac{1}{3}$. Then,
  that in any event, $n_{t+2}\leq \frac{2}{3}n_t$ as claimed.
\end{proof}

From that we conclude that the number of iterations of the \gtx{} algorithm satisfies $K=O(\log
|V_0|)$. We would like to prove a similar result on the edges so that the overall complexity of the
\gtx{} algorithm would be $O(|V_0|\log |V_0|+|E_0|)$. However, this does not follow directly from the
reduction of number of nodes. Indeed we can have a situation where most of the edges of $E_t$ are
between the stars of $G_t$ and thus will remain in the next iteration. Still, we hope that
similarly to the previous proof, this high density of outer edges will lead to strong collapse in
the next iteration.

\iffalse
In the case no singleton are formed during one execution of \extractStar{}, the number of nodes is
reduced by at least a factor $2$ (because each star is made of at least two nodes). On the other hand, we can
construct a graph with singletons where the factor of reduction in number of nodes and edges can be
arbitrarily close to $1$. Consider \autoref{fig:gtx_noreduc}, where $c_1$ and $c_2$ have a degree
$p$ and their neighbors (in light gray) all have a degree $q$, which we can vary from $1$ to $p-1$.
At this step $t$, we thus have $|V_t| = 2(1+p)+pq$ and $|E_t| = 2(p+qp)$. Next we run \extractStar{}
and assume that we first extract the two stars centered in $c_1$ and $c_2$. This reduces the
effective degree of all the middle nodes (in dark gray) to $0$ and they thus form singletons, that
are all connected to the first two stars by \collapseStar{}. As a result, at step $t+1$ we have%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_noreduc_tikz.pdf}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_noreduc}
\end{marginfigure}
$|V_{t+1}| = 2 + pq$ and $|E_{t+1}| = 2pq$, so that
\begin{equation*}
\frac{|V_{t+1}|}{|V_t|} = \frac{pq+2}{p(q+2) + 2} \qquad \text{and} \qquad
\frac{|E_{t+1}|}{|E_t|} = \frac{2pq}{2p(q+1)} = \frac{q}{q+1}
\end{equation*}
When we let $p$ goes to infinity, setting $q=1$ makes $\frac{|V_{t+1}|}{|V_t|}$ tend to
$\nicefrac{1}{3}$ and $\frac{|E_{t+1}|}{|E_t|}$ to $\nicefrac{1}{2}$ while setting $q=p-1$ makes
both $\frac{|V_{t+1}|}{|V_t|}$ and $\frac{|E_{t+1}|}{|E_t|}$ tend to $1$. Note however that in both
cases, this is not really a problem because at the next iteration, there will be only two stars (one
of them will be a singleton). Is it always like that? In other words, can we have a situation where
a large number of singletons is carried over several consecutive iterations?
\fi
